Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-10 paper.tex
Some tweaks

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark}, % VCFShark
most notably the highly-influential Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.

@@

Row-wise storage works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record 
row-wise storage can inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for all rows means retrieving and decompressing \emph{all} 
information from all rows.
The alternative to row-wise storage is \emph{columnar} storage:
instead of grouping toether all the fields for a record,
we group together all the records for a given field.
To our knowledge, the only existing method using columnar 
storage for genetic variation data is 
SeqArray~\citep{zheng2017seqarray} built on 
the GDS container format~\citep{zheng2012high}.

@@

We propose the \texttt{vcfzarr} specification which maps the 
VCF data model into a columnar layout using the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the \texttt{vcfzarr} specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
Note that this approach to storing higher dimensional 
sets Zarr apart from more general columnar formats 
such as Parquet [TODO more details and citations?].
See the Methods for more detail on Zarr and the \texttt{vcfzarr}
specification.

>>>>

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark}, % VCFShark
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.

@@

Row-wise storage works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for all rows means retrieving and decompressing \emph{all} 
information from all rows.
The alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage has become standard [very quick summary of columnar
storage with some refs].
To our knowledge, the only existing method using columnar 
storage for genetic variation data is 
SeqArray~\citep{zheng2017seqarray,zheng2012high}.

@@

We propose the \texttt{vcfzarr} specification which maps the 
VCF data model into a columnar layout using the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the \texttt{vcfzarr} specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
Note that this approach to storing higher dimensional data
sets Zarr apart from more general columnar formats 
such as Parquet [TODO more details and citations?].
See the Methods for more detail on Zarr and the \texttt{vcfzarr}
specification.
